BarnebyLives':' an R package to create herbarium specimen labels and digital data sheets

### 2023-09-17 manuscript/manuscript_APPS.rmd
time trials in, working on results

--- 
title: BarnebyLives an R package to create herbarium specimen labels and digital data sheets
author:  |
    | Reed Clark Benkendorf$^1$^[Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
abstract:  |  
  **Premise:** Depositing specimens to herbaria is a time consuming task. Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated to curatorial tasks and specimen deposition. Despite this, the continual generation of herbaria specimens are essential for research in ecology and evolution. In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks and produce labels.    
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. Census Bureau to provide political jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations. Optionally the package writes driving directions to sites using Google Maps. Finally the package outputs data in a tabular format for review by the user to accept or confirm changes,      
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.     
keywords: |
  Herbarium, collections, natural history museum, geospatial, automation, R, software
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
- \usepackage[export]{adjustbox}
---

@@

@@

\begin{wrapfigure}{l}{0.35\textwidth}
  \centering
    \includegraphics[max size={\textwidth}{\textheight}]{../graphics/plots/flowchart-trim.png}
  \caption{Specified benchmarks at all ESDs which included all three metrics and derived imputed value}
\end{wrapfigure}

@@

```{r load libraries}
library(tidyverse) # data operations
# devtools::install_github('sagesteppe/BarnebyLives', force = TRUE)
library(BarnebyLives) # for helping accession collections
```

## Usage
BarnebyLives is run entirely from within Rstudio. Data may be read in from: Excel, software which can process Comma-separated Values (CSV's) such as LibreOffice, OpenOffice (or Excel), or via the cloud on Googlesheets. The latter two options are documented here and in package vignettes. BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1). Virtually all of the on-disk memory associated with these data are for storing geospatial information, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is available in the package documentation. Functions which require the on-disk data require a path to the data as an argument. 
We anticipate most personal BarnebyLives instances will be less than several gigabytes, and the processing takes relatively little RAM, hence we believe installations can work on hardware as small as Chromebooks, or have the data stored entirely on thumb-drives. The final steps of Barnebylives, generating the labels require working installations of Rmarkdown, a TeX driver (e.g. pdflatex, lualatex, xelatex), and the open source command line tools pdfjam and pdftk. While these steps are run through bash, we have wrapped them in a R functions which bypass the need to enter the commands to a terminal.

@@

@@

### Herbarium Collections
The package was finalized using the primary authors collections from 2023. The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022, *all* of which are un-accessioned. Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used. This results in a data set of XX records for testing, from XX sites located across Western North America FIGURE XX. In total `r n_spp` species (with `r n_sp_authors` authorships), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families were used for testing.

@@

@@

@@

@@

saveRDS(time_directions_grabber, file = 'data/processed/time_directions_grabber')
saveRds(data, file = 'data/processed/data_w_Google_Maps')
```

```{r Total Time of operations}

time_powo_searcher <- readRDS(file = 'data/processed/time_powo_searcher')

time_trials <- data.frame(as.matrix(do.call(rbind, mget(ls(pattern = '^time'))))) %>% 
  rownames_to_column('Function') %>% 
  select(-user.child, -sys.child) %>% 
  mutate(Function = str_remove(Function, 'time_')) %>% 
  mutate(Module = case_when(
    Function %in% c('associate_dropper', 'split_binomials', 'date_parser') ~ 'Style',
    Function %in% c('autofill_checker', 'coords2sf', 'dms2dd', 'physical_grabber', 
                    'political_grabber', 'site_writer') ~ 'Geospatial', 
    Function %in% c('spell_check', 'family_spell_check', 'author_check',  'powo_searcher') ~ 'Taxonomic',
    Function %in% c('time_label_maker', 'time_shipping_manifest', 'time_label_assembly') ~ 'Labels'
  ))

# save these as RDS so operations do not need to be run each time document knits

rm(time_trials)

# rm(ls(pattern = '^time'))
```

It took roughly four (XX exact) minutes to run all local steps of BarnebyLives, and a further XX minutes to search Plants of the World Online, and XX to search Google Maps for site directions.
Nearly all of this time was attributable to the spatial operations (spatial:, taxonomic:, style:) The POWO online search is likely capable of being carried out much faster, xx, if one decreases the pauses (which are by default relatively long) between each query, a tactic used to avoid adversely impacting the website.
The generation of labels is the most time intensive process and consumed around XX minutes for the rendering, XX to combine individual labels four per single sheet of landscape orientated paper, and XX to combine the XX sheets to a single Portable Document Format (PDF).

Even on data which had been manually cleaned and error-checked by a human BarnebyLives was able to reduce transcription errors, identify typos, make nomenclatural suggestions, and reformat text elements for downsteam use.

@@

# AUTHOR CONTRIBUTIONS
The project was conceptualized by R.C.B. The program was written by R.C.B Data collection and analysis were performed by R.C.B. R.C.B. wrote the manuscript with input from all other authors. All authors approved the final version of the manuscript.

@@

\newpage 
\small
# REFERENCES

>>>>

--- 
title: BarnebyLives':' an R package to create herbarium specimen labels and digital data sheets
author:  |
    | Reed Clark Benkendorf$^1$^[Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
abstract:  |  
  **Premise:** Depositing specimens to herbaria is a time consuming task. Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated to curatorial tasks and specimen deposition. Despite this, the continual generation of herbaria specimens are essential for research in ecology and evolution. In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks and produce labels.    
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. Census Bureau to provide political jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations. Optionally the package writes driving directions to sites using Google Maps. Finally the package outputs data in a tabular format for review by the user to accept or confirm changes,      
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.     
keywords: |
  Herbarium, collections, natural history museum, geospatial, automation, R, software
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
- \usepackage[export]{adjustbox}
---

@@

Here we provide a description of the BarnebyLives R package. 
BarnebyLives was named for plant taxonomist extraordinaire Rupert Charles Barneby (1911-200), whom published over 6,500 pages of text, described over 750 taxa, and is notable for balancing his studies at the Willian & Lynda Steere Herbarium at the New York Botanical Garden with annual collection trips in Western North America from 1937-1970, and sporadically until his passing (@welsh2001rupert). 
Select accolades of Rupert include the Asa Gray Award from the American Society of Plant Taxonomists (ASPT) in 1989, the 1991 Engler Silver Medal from the International Association of Plant Taxonomists (IAPT), as well as being one of eight recipients of the International Botanical Congresses's (IBC) Millenium Botany Award (1999) (@welsh2001rupert).

\begin{wrapfigure}{l}{0.35\textwidth}
  \centering
    \includegraphics[max size={\textwidth}{\textheight}]{../graphics/plots/flowchart-trim.png}
  \caption{Recommended workflow.}
\end{wrapfigure}

@@

@@

```{r load libraries}
library(tidyverse) # data operations
# devtools::install_github('sagesteppe/BarnebyLives')
library(BarnebyLives) # for helping accession collections
```

## Usage
BarnebyLives is run entirely from within Rstudio. 
Data may be read in from any common spreadsheet management system or database connection such as Excel, LibreOffice, OpenOffice, or via the cloud on Googlesheets. 
The latter two options are documented here and in package vignettes, detailed descriptions of the required and suggested input columns are located on the Github page (https://github.com/sagesteppe/BarnebyLives *'Input Data Column Names'*) and over 100 real-world examples are on a Google Sheets accessible from the page.
BarnebyLives is atypical of R packages in that it requires a considerable amount of data to operate (Table 1).
Virtually all of the on-disk memory associated with these data are for storing geo-spatial information, setting up a local instance of the program - at whichever scale a user desires (see Figure XX) is available in the package documentation. Functions which require the on-disk data require a path to the data as an argument. 
Manually supplying the argument allows for the users to judiciously decide a storage location suitable for there needs.  
We anticipate most personal BarnebyLives instances will be less than several gigabytes, and the processing takes relatively little RAM, hence we believe installations can work on hardware as small as Chromebooks, or have the data stored entirely on thumb-drives.
The final steps of Barnebylives, generating the labels require working installations of Rmarkdown, a LaTeX installation (e.g. pdflatex, lualatex, xelatex), and the open source command line tools pdfjam and pdftk.
While these steps are run through bash, we have wrapped them in a R functions which bypass the need to enter the commands to a terminal.
Several commands in BarnebyLives require the output from previous functions, and a workflow which satisfies these requirements is presented in FIGURE XX.

@@

# number of collection sites

n_sites <- data %>% 
  distinct(Latitude, Longitude) %>% 
  nrow()

@@

\begin{wrapfigure}{r}{0.35\textwidth}
  \centering
    \includegraphics[width=0.35\textwidth]{../graphics/plots/collections_map-trim.png}
  \caption{The spatial extent (orange), and herbarium collection sites (burgundy) tested in this manuscript.}
\end{wrapfigure}

### Herbarium Collections

The package was finalized using the primary authors collections from 2023. The testing of the package within this manuscript was performed using a subset of their collections from 2018-2022, *all* of which are un-accessioned. Only collections which had identifications to the level of species or lower, and transcribed collection dates and coordinates were used. This results in a data set of `r nrow(data)` records for testing, from `r n_sites` sites located across Western North America FIGURE XX. In total `r n_spp` species (with `r n_sp_authors` authorships), with `r n_infraspecies` infraspecies (`r n_infra_sp_auths` authorships) in `r n_families` families were used for testing.

@@

time_spell_check <- system.time({ # ensure appropriate spellings of the species
  sc_res <- spell_check(data$Binomial, path = p2tax)
})

@@

data$Family == data$Family

@@

rm(p2geo, p2tax)
```

@@

saveRDS(time_directions_grabber, file = 'data/processed/time_directions_grabber')
saveRDS(data, file = 'data/processed/data_w_Google_Maps')
```

```{r Total Time of operations, eval = F}

time_powo_searcher <- readRDS(file = 'data/processed/time_powo_searcher')
time_directions_grabber <- readRDS('data/processed/time_directions_grabber')

time_trials <- data.frame(as.matrix(do.call(rbind, mget(ls(pattern = '^time'))))) %>% 
  rownames_to_column('Function') %>% 
  select(-user.child, -sys.child) %>% 
  mutate(Function = str_remove(Function, 'time_')) %>% 
  mutate(Module = case_when(
    Function %in% c('associate_dropper', 'split_binomials', 'date_parser') ~ 'Style',
    Function %in% c('autofill_checker', 'coords2sf', 'dms2dd', 'physical_grabber', 
                    'political_grabber', 'site_writer', 'geodata_writer', 'directions_grabber') ~ 'Geospatial', 
    Function %in% c('spell_check', 'family_spell_check', 'author_check',  'powo_searcher') ~ 'Taxonomic',
    Function %in% c('time_label_maker', 'time_shipping_manifest', 'time_label_assembly') ~ 'Labels'
    ),
    Group = if_else(Function %in% c('powo_searcher', 'directions_grabber'), 'Online', 'Local')
  )

# save these as RDS so operations do not need to be run each time document knits
saveRDS(time_trials, file = 'data/processed/time_trials')

rm(time_trials)
rm(list=ls(pattern = 'time'))
```

```{r load in time trials}
time_trials <- readRDS('data/processed/time_trials')
tt_local <- time_trials[time_trials$Group=='Local',]
```

It took roughly three minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) minutes to run all local steps of BarnebyLives, and a further 15 minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']` to search Google Maps and write directions to sites.
Nearly all of the local run time is attributable to the spatial operations (spatial: `r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`s, style: `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`, taxonomic: `r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`). 
The generation of labels is the most time intensive process and consumed around XX minutes for the rendering, XX to combine individual labels four per single sheet of landscape orientated paper, and XX to combine the XX sheets to a single Portable Document Format (PDF).

```{r}
rm(time_trials, tt_local)
```

## Results

```{r import results}
data_p <- readRDS(file = 'data/processed/data_w_POWO_search')

drop_na(data_p, Binomial_authority) %>% 
  nrow() # nearly all relating to the absence/presence of a period

drop_na(data_p, Infra_auth_issues) %>% 
  nrow() # nearly all relating to the absence/presence of a period

drop_na(data_p, Long_AutoFill_Flag) %>% 
  nrow() # 4 auto fill cases

drop_na(data_p, Lat_AutoFill_Flag) %>% 
  nrow() # 2 autofill cases

## Family spell check; this is a comparision of my input spelling, and the results
## of the family spell check function

Family_in_positions <- data %>% 
  drop_na(Family)
Family_in <- Family_in_positions %>% 
  filter(Collection_number %in% data_p$Collection_number) %>% 
  arrange(Collection_number) %>% 
  pull(Family)

Family_out <- data_p %>% 
  filter(Collection_number %in% Family_in_positions$Collection_number) %>% 
  pull(Family)
Family_in[ which( Family_in != Family_out ) ]

## Family Level Results
# these species were found by POW, and according to POW, the species are in the f
# following families. 
data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(POW_Family != 'NOT FOUND') %>% 
  filter(POW_Family != Family) %>% 
  select(Collection_number, Full_name, Family, POW_Family) %>% 
  mutate(Situation = 
    case_when(
      Family == 'Hydrophyllaceae' ~ 'Preferred', 
      Family %in% c('Plantaginaceae', 'Athyriaceae') ~ 'Outdated',
      Family %in% c('Saxifragaceae', 'Solanaceae') ~ 'Submitter_Incorrect', 
      Family == 'Bataceae' ~ 'Internal Error', 
      .default = 'Typo'
    ))

data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(POW_Genus == 'NOT FOUND') %>% 
  select(Collection_number, Full_name, Genus, POW_Genus)

# how many of my genera were mis-spelled??
data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(POW_Genus != 'NOT FOUND') %>% 
  filter(POW_Genus != Genus)

# the mismatch between POW genus and not
data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(POW_Genus != 'NOT FOUND') %>% 
  filter(POW_Genus != Genus) %>% 
  select(Collection_number, Full_name, Genus, POW_Genus) %>% 
  mutate(Situation = 
    case_when(
      Genus %in% c('Pleuraphis', 'Glandularia') ~ 'Preferred', # check Pleuraphis, i am confused on this still....
      Genus %in% c('Oenothera', 'Cryptantha', 'Ivesia', 'Lotus', 'Polygala', 'Minuartia',
                    'Pseudostellaria', 'Achnatherum', 'Scirpus', 'Peritoma', 'Arenaria', 
                    'Spartina', 'Polygonum', 'Atriplex', 'Cymopterus', 'Trisetum', 'Bahia', 
                   'Oryzopsis', 'Potentilla', 'Pascopyrum'
                    ) ~ 'Outdated',
      Genus %in% c('Vaccinium', 'Eleocharis') ~ 'Internal Error',
      .default = 'Typo'
    ),
    Situation = if_else(Full_name == 'Oenothera caespitosa Nutt.', 'Internal Error', Situation))

data_p %>% 
  filter(Binomial_authority != POW_Name_authority)

data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(Epithet != POW_Epithet ) %>% 
  select(Collection_number, Full_name, Epithet, POW_Epithet)

# NEED TO WORK ON INPUT DATA, INCONSISTENT USE OF AUTHORSHIP IN THE FIRST COLUMN
data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(Full_name != POW_Full_name) %>% 
  select(Collection_number, Full_name, Binomial, POW_Full_name) 
```

Even on data which had been manually cleaned and error-checked by a human several times BarnebyLives was able to reduce transcription errors, identify typos, make nomenclature suggestions, and reformat text elements for downstream use. The number of family misspellings were XX (% percent), the number of misspelled genera were XX (% percent), the number of misspelled binomials were XX (% percent). The number of author abbreviations which were not in the appropriate format were XX (% percent), generally the presence or absence of a period were the issue. Plants of the World Online was able to identify XX new names for the submitted taxa, XX of which the author adopted. XX records were appropriately flagged for issues with auto fill incrementation of the longitude value, and three of these records were also auto-flagged for increases in latitude values (% of records).

@@

# AUTHOR CONTRIBUTIONS
The project was conceptualized by R.C.B. The program was written by R.C.B. Data collection and analysis were performed by R.C.B. R.C.B. wrote the manuscript with input from all other authors. All authors approved the final version of the manuscript.

@@

\small
# REFERENCES

<div id="refs"></div>

# SUPPORTING INFORMATION
Additional supporting information can be found online in the Supporting Information section at the end of this article.

**Appendix S1.** A table of all time trials for each function.
